Understanding Structure and Selectivity in Nanostructured Lyotropic Liquid Crystal Membranes using Molecular Simulations

### 2020-04-05 Ben_Manuscripts/thesis/transport.tex
Updates to outline and minor changes to figures

  %\graphicspath{{./figs/}}
  %\graphicspath{{/home/bcoscia/PycharmProjects/LLC_Membranes/Ben_Manuscripts/transport/figures}}

@@

  Membranes capable of separating nm-sized solutes with high selectivity and
  permeability are highly desirable for a number of applications. For example,
  separation of salt from seawater or other briny sources can yield potable
  drinking water.~\cite{fritzmann_state---art_2007} Removal of organic
  micropollutants, such as personal care products, pesticides and
  pharmaceuticals, from surface and groundwaters, can have large benefits for
  public health.~\cite{schwarzenbach_challenge_2006} Finally, one can purify and
  recover dissolved species present in complex hydraulic fracturing flowback
  water waste streams which would help mitigate the effects of deep well
  injection and can be sold for profit.~\cite{dischinger_application_2017}

  All of these separations are possible in part with current commercial
  membrane separation techniques, but all still suffer from serious drawbacks.
  Currently, reverse osmosis (RO) and nanofiltration (NF) dominate commercial
  membrane separations of small molecules.~\cite{warsinger_review_2018} RO
  membranes are typically dense polymer matrices that separate solutes based on
  differences in their solubility and permeability in the membrane
  material.~\cite{fritzmann_state---art_2007} Although they can perform highly
  selective separations, high feed pressures and thus large amounts of energy are
  necessary in order to generate a useful permeate
  flux.~\cite{van_der_bruggen_review_2003} NF membranes have well-defined
  nm-sized pores which can give the same permeate flux as RO with lower applied
  pressure.~\cite{hilal_comprehensive_2004} However, the pores are not uniform in
  size, limiting their selectivity.~\cite{werber_materials_2016}.

  Under the right conditions, amphiphilic molecules can self-assemble into
  ordered nanostructures that may be capable of performing highly selective
  separations with permeabilities comparable to
  NF.~\cite{alexandridis_amphiphilic_2000} The uniform-sized aqueous pores
  characteristic of these types of systems enforce a molecular-size cut-off,
  eliminating issues related to the polydisperity of NF
  pores.~\cite{zhou_supported_2005} In addition to size-exclusion, it may be
  possible to control the chemical environment within the nanopores by altering
  the identities of the hydrophilic moieties that form the pores, affording
  solute-specific separations.~\cite{dischinger_effect_2017}

  Amphiphilic lyotropic liquid crystals (LLCs) are known to form the type I
  bicontinuous cubic (Q\textsubscript{I}) and inverted hexagonal (H\textsubscript{II})
  phases.~\cite{smith_ordered_1997,dischinger_application_2017,carter_glycerol-based_2012,hatakeyama_nanoporous_2010}
  When cross-linked, these materials can function as mechanically strong
  separations membranes.~\cite{zhou_supported_2005} The H\textsubscript{II} phase
  has hexagonally packed, straight pores which is the ideal geometry for high
  throughtput separations. However, the synthesis of H\textsubscript{II}
  membranes with highly aligned hexagonal mesophases has been a challenge that
  has been met with limited success and has prevented the material's viability as
  a separations membrane.~\cite{feng_scalable_2014,feng_thin_2016} The
  Q\textsubscript{I} phase consists of a three dimensional network of tortuous
  and interconnected pores.  It's more facile synthesis has yielded it more
  success. The latest Q\textsubscript{I} phase membranes have been shown to
  exhibit selectivities higher than commercial NF and RO membranes with permeabilities
  higher than commercial RO membranes.~\cite{dischinger_application_2017}

  Although Q\textsubscript{I} membranes have shown promising results, it
  is not clear how to controllably design them for specific separations
  beyond what has already been achieved. Dischinger et al. found that they
  do not separate purely on the basis of size. Empirical modeling revealed
  correlations between various solute physicochemical properties and 
  observed solute rejection. Although there was some agreement with their
  empirical model, it does not offer a sufficiently detailed explanation
  of governing molecular interactions.~\cite{dischinger_effect_2017}

@@

  Python scripts used to set up systems and conduct post-simulation trajectory 
  analysis (as described below and in the Supporting Information) are available 
  online at \\
  \texttt{https://github.com/shirtsgroup/LLC\_Membranes}. The 
  appropriate scripts to use for the subsequent calculations are summarized in 
  Table~\ref{S-table:python_scripts_transport} of the Supporting Information. We ran all 
  MD simulations and energy minimizations using GROMACS 2018.3.~\cite{bekker_gromacs:_1993,berendsen_gromacs:_1995,van_der_spoel_gromacs:_2005,hess_gromacs_2008}

@@

  We initially attempted to equilibrate our system with water by allowing water
  molecules to naturally penetrate the membrane from a water bath separating
  periodic images of the system in the $z$-direction. We allowed a dry, previously
  equilibrated system to further equilibrate in coexistence with a 3 nm-thick 
  (in the $z$-direction) layer of water. Although the water appears to equilibrate
  after 1000 ns of simulation, the density of water along the pore axis
  is close to zero at the membrane center, implying that there is a kinetic
  barrier to water filling the pores which we will be unable to overcome using
  conventional simulations. See Section~\ref{S-section:water_content_equil} 
  of the Supporting Information for an expanded explanation and supporting figures.

@@

  We removed the water reservoir and allowed the pore and tail water contents
  to equilibrate with 5 and 10 wt\% total water with a 3:2 ratio of pore to tail
  water. We considered the water content equilibrated once the number of water
  molecules in each region plateaued. The 5 wt\% system did not plateau until
  $\sim$ 600 ns while the 10 wt \% water system equilibrated within the first 100
  ns of simulation (see Figure~\ref{S-fig:5wt_offset_equil} and
  Figure~\ref{S-fig:10wt_offset_equil} of the Supporting Information). The
  equilibrated pores contain 72 \% and 69 \% of the total water in the 5 and 10
  wt \% systems respectively. We cross-linked the equilibrated solvated systems,
  then allowed them to equilibrate further for 100 ns. The water contents in
  each region does not change significantly in either case (see
  Figure~\ref{S-fig:xlinked_solvation_equilibration} of the Supporting
  Information). The `wet' equilibration and cross-linking procedures follow the
  methods outlined in our previous work.~\cite{coscia_understanding_2019}

@@

  To study a given solute, we added 6 solute molecules, equally spaced in $z$, to the
  center of each pore of the equilibrated cross-linked configuration, for a total of 24
  solute molecules. We chose 6 solutes per pore to provide a balance of a useful amount of 
  data for generating statistics and a low degree of interaction between solutes (see 
  Section~\ref{S-section:solute_interaction} of the Supporting Information). At each
  insertion point we placed a randomly oriented solute molecule then ran a short 
  energy minimization. We allowed the solutes to equilibrate at 300K for 5 ns using Berendsen
  pressure control then collected transport data over the course of 1 $\mu$s MD simulations
  with pressure controlled by the Parrinello-Rahman barostat.

@@

  We fixed the length of each simulated trajectory so that we could compare the total
  MSD between different solutes without the influence of the ageing phenomenon.
  In systems such as ours where solutes show hopping behavior between long periods 
  of immobility, ageing is defined by the tendency of the average slope of an MSD 
  curve to decrease as the length of trajectories are increased~\cite{metzler_anomalous_2014}.
  Since the maximum measured dwell time can be no longer than the total length of a simulated
  trajectory, longer dwell times are incorporated into the calculation as measurement
  time or trajectory length is increased, lowering the average MSD. Because the solute 
  MSDs are non-linear and because of the ageing phenomenon, we did not attempt to calculate
  a diffusion constant as one might for a Brownian particle with a linear MSD. Instead,
  the reported MSD values represent the average MSD for a given	solute after a 400
  ns time lag. Our results are shown to be insensitive to our choice of time lag in 
  Section~\ref{S-section:lag_sensitivity} of the Supporting Information.

@@

  We differentiated between the length of hops inside and outside of the 
  pores. For 10 wt\% water systems, we consider a solute to be in the pore
  if it is within 0.75 nm of a pore center. This radial cut-off maximizes 
  the difference between average hop lengths in and out of the pore. 
  See Section~\ref{S-section:pore-region} of the Supporting
  Information for further details on this optimization.

@@

  Based on the geometric criteria proposed by Luzar and Chandler 
  \cite{luzar_effect_1996}, we determined a hydrogen bond to exist if the
  distance between the donor, D, and acceptor, A, atoms is less than 
  3.5 \AA~and the angle formed by D--H$\cdot\cdot\cdot$A is less than 30\degree. Attempts
  to describe a hydrogen bond in the context of molecular simulations has
  yielded a number of definitions with no true consensus 
  \cite{prada-gracia_quest_2013} especially since the geometry of hydrogen
  bonds has some dependence on the system being studied. The definition of
  Luzar and Chandler is easily visualized for trajectories using the 
  \texttt{hbonds} representation of the Visual Molecular Dynamics (VMD) software 
  package which allows us to directly check the validity of identified hydrogen 
  bonds~\cite{humphrey_VMD:_1996}. In Section~\ref{S-section:hbond_sensitivity} 
  of the Supporting Information, we show that our conclusions are insensitive
  to this definition within a reasonable range of distances and angles.

@@

  Hydrogen bond lifetimes appear to be distributed according to a power law or an
  exponential function (see Figure~\ref{S-fig:lifetime_distributions} of the 
  Supporting Information). A number of researchers provide evidence that supports a 
  power law distribution~\cite{starr_fast_1999,martiniano_insights_2013}. However, these
  studies were done on extremely short timescales relative to ours, outputting positions
  every time step. Voloshin et al. studied hydrogen bonding on multiple timescales and 
  observed exponential behavior on the longest timescales~\cite{voloshin_hydrogen_2009}. 
  Due to memory limitations, we could not collect data frequently enough to provide a
  sufficient answer to this question so when comparing solutes, we use the 95\textsuperscript{th} 
  percentile of hydrogen bond dwell times. This places emphasis on solutes with 
  long dwell times. We reported the standard error of this calculation by bootstrapping
  the distribution of dwell times.

@@

  Although the pores are often described as straight, they have a
  small degree of tortuosity which disrupts the RDF calculation. We 
  tried to mitigate the effects of tortuosity by calculating the RDF
  with respect to splines that run through the pore centers. See
  Section~\ref{S-section:splines} of the Supporting Information for 
  a graphical illustration. Each spline consists of 10 points, equally
  spaced in the $z$-direction, whose ($x$, $y$) coordinates are defined
  based on the COM of all head groups closest, in $z$, to the
  given point. When calculating RDFs, the radial distance from the pore 
  center is based on the distance between the solute COM and
  the linearly interpolated ($x$, $y$) coordinates of the pore center 
  calculated based on the closest two spline points. Using the splines, 
  we calculated the tortuosity of the pores by calculating the ratio 
  $\dfrac{L}{Z}$ where $L$ is the length of the spline and $Z$ is the 
  length of the unit cell in the $z$-direction. The average tortuosity
  of each pore is 1.03 $\pm$ 0.01 and 1.07 $\pm$ 0.02 in the 5 and 10 wt\%
  water systems respectively.

@@

  There is an appreciable amount of water that partitions into the distal tail
  region of both systems. 28\% and 31\% of the total water is present in the
  distal tail regions of the 5 and 10 wt\% water systems respectively. See 
  Section~\ref{S-section:water_content_equil} of the Supporting Information 
  for more details on the equilibrations that led to this conclusion. The partition
  is due to a combination of the distal tail region's lower density (see 
  Figure~\ref{fig:total_density}) as well as oxygen atoms at the ends of each
  monomer tail which can further stabilize water molecules.

>>>>

@@

%  Membranes capable of separating nm-sized solutes with high selectivity and
%  permeability are highly desirable for a number of applications. For example,
%  separation of salt from seawater or other briny sources can yield potable
%  drinking water.~\cite{fritzmann_state---art_2007} Removal of organic
%  micropollutants, such as personal care products, pesticides and
%  pharmaceuticals, from surface and groundwaters, can have large benefits for
%  public health.~\cite{schwarzenbach_challenge_2006} Finally, one can purify and
%  recover dissolved species present in complex hydraulic fracturing flowback
%  water waste streams which would help mitigate the effects of deep well
%  injection and can be sold for profit.~\cite{dischinger_application_2017} 
%
%  All of these separations are possible in part with current commercial
%  membrane separation techniques, but all still suffer from serious drawbacks.
%  Currently, reverse osmosis (RO) and nanofiltration (NF) dominate commercial
%  membrane separations of small molecules.~\cite{warsinger_review_2018} RO
%  membranes are typically dense polymer matrices that separate solutes based on
%  differences in their solubility and permeability in the membrane
%  material.~\cite{fritzmann_state---art_2007} Although they can perform highly
%  selective separations, high feed pressures and thus large amounts of energy are
%  necessary in order to generate a useful permeate
%  flux.~\cite{van_der_bruggen_review_2003} NF membranes have well-defined
%  nm-sized pores which can give the same permeate flux as RO with lower applied
%  pressure.~\cite{hilal_comprehensive_2004} However, the pores are not uniform in
%  size, limiting their selectivity.~\cite{werber_materials_2016}.
%
%  Under the right conditions, amphiphilic molecules can self-assemble into
%  ordered nanostructures that may be capable of performing highly selective
%  separations with permeabilities comparable to
%  NF.~\cite{alexandridis_amphiphilic_2000} The uniform-sized aqueous pores
%  characteristic of these types of systems enforce a molecular-size cut-off,
%  eliminating issues related to the polydisperity of NF
%  pores.~\cite{zhou_supported_2005} In addition to size-exclusion, it may be
%  possible to control the chemical environment within the nanopores by altering
%  the identities of the hydrophilic moieties that form the pores, affording
%  solute-specific separations.~\cite{dischinger_effect_2017}
%
%  Amphiphilic lyotropic liquid crystals (LLCs) are known to form the type I
%  bicontinuous cubic (Q\textsubscript{I}) and inverted hexagonal (H\textsubscript{II})
%  phases.~\cite{smith_ordered_1997,dischinger_application_2017,carter_glycerol-based_2012,hatakeyama_nanoporous_2010}
%  When cross-linked, these materials can function as mechanically strong
%  separations membranes.~\cite{zhou_supported_2005} The H\textsubscript{II} phase
%  has hexagonally packed, straight pores which is the ideal geometry for high
%  throughtput separations. However, the synthesis of H\textsubscript{II}
%  membranes with highly aligned hexagonal mesophases has been a challenge that
%  has been met with limited success and has prevented the material's viability as
%  a separations membrane.~\cite{feng_scalable_2014,feng_thin_2016} The
%  Q\textsubscript{I} phase consists of a three dimensional network of tortuous
%  and interconnected pores.  It's more facile synthesis has yielded it more
%  success. The latest Q\textsubscript{I} phase membranes have been shown to
%  exhibit selectivities higher than commercial NF and RO membranes with permeabilities
%  higher than commercial RO membranes.~\cite{dischinger_application_2017}
%
%  Although Q\textsubscript{I} membranes have shown promising results, it
%  is not clear how to controllably design them for specific separations
%  beyond what has already been achieved. Dischinger et al. found that they
%  do not separate purely on the basis of size. Empirical modeling revealed
%  correlations between various solute physicochemical properties and 
%  observed solute rejection. Although there was some agreement with their
%  empirical model, it does not offer a sufficiently detailed explanation
%  of governing molecular interactions.~\cite{dischinger_effect_2017}

@@

  Python scripts used to set up systems and conduct post-simulation trajectory 
  analysis (as described below and in Appendix B) are available 
  online at \\
  \texttt{https://github.com/shirtsgroup/LLC\_Membranes}. The 
  appropriate scripts to use for the subsequent calculations are summarized in 
  Table~\ref{S-table:python_scripts_transport} of Appendix B. We ran all 
  MD simulations and energy minimizations using GROMACS 2018.3.~\cite{bekker_gromacs:_1993,berendsen_gromacs:_1995,van_der_spoel_gromacs:_2005,hess_gromacs_2008}

@@

  We initially attempted to equilibrate our system with water by allowing water
  molecules to naturally penetrate the membrane from a water bath separating
  periodic images of the system in the $z$-direction. We allowed a dry, previously
  equilibrated system to further equilibrate in coexistence with a 3 nm-thick 
  (in the $z$-direction) layer of water. Although the water appears to equilibrate
  after 1000 ns of simulation, the density of water along the pore axis
  is close to zero at the membrane center, implying that there is a kinetic
  barrier to water filling the pores which we will be unable to overcome using
  conventional simulations. See Section~\ref{S-section:water_content_equil} 
  of Appendix B for an expanded explanation and supporting figures.

@@

  We removed the water reservoir and allowed the pore and tail water contents
  to equilibrate with 5 and 10 wt\% total water with a 3:2 ratio of pore to tail
  water. We considered the water content equilibrated once the number of water
  molecules in each region plateaued. The 5 wt\% system did not plateau until
  $\sim$ 600 ns while the 10 wt \% water system equilibrated within the first 100
  ns of simulation (see Figure~\ref{S-fig:5wt_offset_equil} and
  Figure~\ref{S-fig:10wt_offset_equil} of Appendix B). The
  equilibrated pores contain 72 \% and 69 \% of the total water in the 5 and 10
  wt \% systems respectively. We cross-linked the equilibrated solvated systems,
  then allowed them to equilibrate further for 100 ns. The water contents in
  each region does not change significantly in either case (see
  Figure~\ref{S-fig:xlinked_solvation_equilibration} of Appendix B). The `wet' 
  equilibration and cross-linking procedures follow the
  methods outlined in our previous work.~\cite{coscia_understanding_2019}

@@

  To study a given solute, we added 6 solute molecules, equally spaced in $z$, to the
  center of each pore of the equilibrated cross-linked configuration, for a total of 24
  solute molecules. We chose 6 solutes per pore to provide a balance of a useful amount of 
  data for generating statistics and a low degree of interaction between solutes (see 
  Section~\ref{S-section:solute_interaction} of Appendix B). At each
  insertion point we placed a randomly oriented solute molecule then ran a short 
  energy minimization. We allowed the solutes to equilibrate at 300K for 5 ns using Berendsen
  pressure control then collected transport data over the course of 1 $\mu$s MD simulations
  with pressure controlled by the Parrinello-Rahman barostat.

@@

  We fixed the length of each simulated trajectory so that we could compare the total
  MSD between different solutes without the influence of the ageing phenomenon.
  In systems such as ours where solutes show hopping behavior between long periods 
  of immobility, ageing is defined by the tendency of the average slope of an MSD 
  curve to decrease as the length of trajectories are increased~\cite{metzler_anomalous_2014}.
  Since the maximum measured dwell time can be no longer than the total length of a simulated
  trajectory, longer dwell times are incorporated into the calculation as measurement
  time or trajectory length is increased, lowering the average MSD. Because the solute 
  MSDs are non-linear and because of the ageing phenomenon, we did not attempt to calculate
  a diffusion constant as one might for a Brownian particle with a linear MSD. Instead,
  the reported MSD values represent the average MSD for a given	solute after a 400
  ns time lag. Our results are shown to be insensitive to our choice of time lag in 
  Section~\ref{S-section:lag_sensitivity} of Appendix B.

@@

  We differentiated between the length of hops inside and outside of the 
  pores. For 10 wt\% water systems, we consider a solute to be in the pore
  if it is within 0.75 nm of a pore center. This radial cut-off maximizes 
  the difference between average hop lengths in and out of the pore. 
  See Section~\ref{S-section:pore-region} of Appendix B for further details on this optimization.

@@

  Based on the geometric criteria proposed by Luzar and Chandler 
  \cite{luzar_effect_1996}, we determined a hydrogen bond to exist if the
  distance between the donor, D, and acceptor, A, atoms is less than 
  3.5 \AA~and the angle formed by D--H$\cdot\cdot\cdot$A is less than 30\degree. Attempts
  to describe a hydrogen bond in the context of molecular simulations has
  yielded a number of definitions with no true consensus 
  \cite{prada-gracia_quest_2013} especially since the geometry of hydrogen
  bonds has some dependence on the system being studied. The definition of
  Luzar and Chandler is easily visualized for trajectories using the 
  \texttt{hbonds} representation of the Visual Molecular Dynamics (VMD) software 
  package which allows us to directly check the validity of identified hydrogen 
  bonds~\cite{humphrey_VMD:_1996}. In Section~\ref{S-section:hbond_sensitivity} 
  of Appendix B, we show that our conclusions are insensitive
  to this definition within a reasonable range of distances and angles.

@@

  Hydrogen bond lifetimes appear to be distributed according to a power law or an
  exponential function (see Figure~\ref{S-fig:lifetime_distributions} of
  Appendix B). A number of researchers provide evidence that supports a 
  power law distribution~\cite{starr_fast_1999,martiniano_insights_2013}. However, these
  studies were done on extremely short timescales relative to ours, outputting positions
  every time step. Voloshin et al. studied hydrogen bonding on multiple timescales and 
  observed exponential behavior on the longest timescales~\cite{voloshin_hydrogen_2009}. 
  Due to memory limitations, we could not collect data frequently enough to provide a
  sufficient answer to this question so when comparing solutes, we use the 95\textsuperscript{th} 
  percentile of hydrogen bond dwell times. This places emphasis on solutes with 
  long dwell times. We reported the standard error of this calculation by bootstrapping
  the distribution of dwell times.

@@

  Although the pores are often described as straight, they have a
  small degree of tortuosity which disrupts the RDF calculation. We 
  tried to mitigate the effects of tortuosity by calculating the RDF
  with respect to splines that run through the pore centers. See
  Section~\ref{S-section:splines} of Appendix B for 
  a graphical illustration. Each spline consists of 10 points, equally
  spaced in the $z$-direction, whose ($x$, $y$) coordinates are defined
  based on the COM of all head groups closest, in $z$, to the
  given point. When calculating RDFs, the radial distance from the pore 
  center is based on the distance between the solute COM and
  the linearly interpolated ($x$, $y$) coordinates of the pore center 
  calculated based on the closest two spline points. Using the splines, 
  we calculated the tortuosity of the pores by calculating the ratio 
  $\dfrac{L}{Z}$ where $L$ is the length of the spline and $Z$ is the 
  length of the unit cell in the $z$-direction. The average tortuosity
  of each pore is 1.03 $\pm$ 0.01 and 1.07 $\pm$ 0.02 in the 5 and 10 wt\%
  water systems respectively.

@@

  There is an appreciable amount of water that partitions into the distal tail
  region of both systems. 28\% and 31\% of the total water is present in the
  distal tail regions of the 5 and 10 wt\% water systems respectively. See 
  Section~\ref{S-section:water_content_equil} of Appendix B 
  for more details on the equilibrations that led to this conclusion. The partition
  is due to a combination of the distal tail region's lower density (see 
  Figure~\ref{fig:total_density}) as well as oxygen atoms at the ends of each
  monomer tail which can further stabilize water molecules.
